---
title: "Analyzing Student Academic Performance"
subtitle: "Factors Influencing Exam Scores"
author: "Satyam Kumar"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
```

# Introduction

## Problem Statement

Educational institutions want to understand which background and preparation
factors are associated with how well students perform in exams. This report
analyzes the **"Students Performance in Exams"** Dataset, which records the
math, reading, and writing scores of 1,000 students along with demographic
and preparatory attributes such as gender, parental level of education,
lunch type (a proxy for socio-economic status), and whether the student
completed a test-preparation course.

**Objectives of this project:**

1. Import and inspect the raw dataset.
2. Clean and preprocess the data for analysis.
3. Explore the data using summary statistics and visualizations.
4. Test whether specific factors (test preparation, lunch type, gender) are
   statistically associated with exam performance.
5. Summarize findings and draw conclusions.

## Loading Required Packages

We use `dplyr` and `tidyr` for data manipulation and `ggplot2` for
visualization. These are the standard "tidyverse" tools for data analysis
in R.

```{r libraries}
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Data Importing

The dataset is provided as a CSV file. We import it using `read.csv()`,
which is R's base function for reading comma-separated value files into a
data frame — the core rectangular data structure R uses for tabular data.

```{r import}
students <- read.csv("StudentsPerformance.csv", stringsAsFactors = FALSE)

# Basic structure: number of rows/columns and data types of each column
str(students)
```

```{r import-head}
# First 6 rows, to visually confirm the import looks correct
head(students)
```

```{r import-dim}
cat("Rows:", nrow(students), " Columns:", ncol(students), "\n")
```

**Interpretation:** The dataset has 1,000 rows (one per student) and 8
columns: 5 categorical attributes (`gender`, `race/ethnicity`,
`parental level of education`, `lunch`, `test preparation course`) and 3
numeric exam scores (`math score`, `reading score`, `writing score`).

# Data Preprocessing and Cleaning

Raw data is rarely analysis-ready. In this section we standardize column
names, check for missing values and duplicate rows, fix data types, and
engineer a couple of useful derived variables.

## Standardizing Column Names

The original column names contain spaces and slashes, which are awkward to
reference in R code (they must be wrapped in backticks). We rename them to
clean, `snake_case` names using `rename()`.

```{r rename}
students <- students %>%
  rename(
    gender          = gender,
    ethnicity_group = `race.ethnicity`,
    parent_education = `parental.level.of.education`,
    lunch           = lunch,
    test_prep       = `test.preparation.course`,
    math_score      = `math.score`,
    reading_score   = `reading.score`,
    writing_score   = `writing.score`
  )

colnames(students)
```

*Note:* `read.csv()` automatically converts spaces and slashes in the
original header row into dots (e.g. `race/ethnicity` becomes
`race.ethnicity`), so we reference those dotted names on the right-hand
side of `rename()`.

## Checking for Missing Values

```{r missing}
colSums(is.na(students))
```

**Interpretation:** There are no missing values in any column, so no
imputation or row-removal is required for this step. We still show this
check explicitly because verifying data completeness is a mandatory part
of any cleaning pipeline, even when the result is "no problem found."

## Checking for Duplicate Rows

```{r duplicates}
sum(duplicated(students))
```

No duplicate rows are present.

## Fixing Data Types

Categorical columns are currently stored as plain text (`character`). We
convert them to R's `factor` type, which is the appropriate type for
categorical variables and is required for grouped statistics, plotting
order, and modeling.

```{r factors}
students <- students %>%
  mutate(
    gender            = factor(gender),
    ethnicity_group   = factor(ethnicity_group),
    parent_education  = factor(parent_education,
                                levels = c("some high school", "high school",
                                           "some college", "associate's degree",
                                           "bachelor's degree", "master's degree")),
    lunch             = factor(lunch),
    test_prep         = factor(test_prep, levels = c("none", "completed"))
  )

str(students)
```

## Feature Engineering

We add two derived variables that make later analysis easier:

* `average_score` — the mean of the three subject scores, used as a single
  overall-performance metric.
* `pass_fail` — a simple pass/fail label using 60 as an illustrative
  passing threshold on the average score.

```{r feature-engineering}
students <- students %>%
  mutate(
    average_score = round((math_score + reading_score + writing_score) / 3, 1),
    pass_fail = factor(ifelse(average_score >= 60, "Pass", "Fail"),
                        levels = c("Pass", "Fail"))
  )

head(students %>% select(gender, test_prep, math_score, reading_score,
                          writing_score, average_score, pass_fail))
```

The cleaned dataset now has `r ncol(students)` columns and `r nrow(students)`
rows, with correct data types and two new analysis-ready variables.

# Data Analysis and Statistical Exploration

## Overall Summary Statistics

```{r summary}
summary(students %>% select(math_score, reading_score, writing_score, average_score))
```

**Interpretation:** Average scores across all three subjects cluster
around the high-60s, with reading scores tending to be slightly higher on
average than math scores.

## Performance by Test Preparation Course

```{r group-test-prep}
students %>%
  group_by(test_prep) %>%
  summarise(
    n = n(),
    mean_math    = round(mean(math_score), 1),
    mean_reading = round(mean(reading_score), 1),
    mean_writing = round(mean(writing_score), 1),
    mean_average = round(mean(average_score), 1)
  )
```

**Interpretation:** Students who completed the test preparation course
score higher on average across all three subjects than those who did not,
suggesting a positive association between preparation and performance.

## Performance by Lunch Type (Socio-Economic Proxy)

```{r group-lunch}
students %>%
  group_by(lunch) %>%
  summarise(
    n = n(),
    mean_average = round(mean(average_score), 1),
    sd_average   = round(sd(average_score), 1)
  )
```

**Interpretation:** Students receiving a "standard" lunch (generally
associated with higher household income) show a noticeably higher average
score than those on "free/reduced" lunch, consistent with prior education
research linking socio-economic status to academic outcomes.

## Correlation Between Subject Scores

```{r correlation}
cor(students %>% select(math_score, reading_score, writing_score))
```

**Interpretation:** All three subject scores are strongly positively
correlated (values above 0.8), meaning students who score well in one
subject tend to score well in the others — an overall "academic ability"
factor appears to be at play rather than subject-specific skills acting
independently.

## Statistical Test: Does Test Preparation Significantly Affect Scores?

We use an independent-samples **t-test** to check whether the difference
in average scores between students who completed the test-prep course and
those who did not is statistically significant (not just due to random
sampling variation).

```{r t-test}
t.test(average_score ~ test_prep, data = students)
```

**Interpretation:** The p-value is well below 0.05, so we reject the null
hypothesis of no difference. Completing the test preparation course is
associated with a statistically significant increase in average exam
score.

## Statistical Test: Does Gender Affect Math Scores?

```{r anova-gender}
t.test(math_score ~ gender, data = students)
```

**Interpretation:** Compare the p-value to the 0.05 significance
threshold: if it is below 0.05, there is a statistically significant
difference in math scores between genders in this sample.

## Simple Linear Regression: Predicting Writing Score from Reading Score

```{r regression}
model <- lm(writing_score ~ reading_score, data = students)
summary(model)
```

**Interpretation:** The coefficient on `reading_score` shows how many
points writing score is predicted to change for every one-point increase
in reading score. The R-squared value indicates the proportion of
variance in writing scores explained by reading scores alone — a high
value here confirms the strong correlation seen earlier.

# Graphs and Plots

## Distribution of Average Scores

```{r plot-histogram}
ggplot(students, aes(x = average_score)) +
  geom_histogram(binwidth = 5, fill = "#4C72B0", color = "white") +
  labs(title = "Distribution of Average Student Scores",
       x = "Average Score", y = "Number of Students") +
  theme_minimal()
```

**Interpretation:** The distribution is roughly bell-shaped and centered
around the high-60s, consistent with the summary statistics computed
earlier, with relatively few students at the extreme low or high ends.

## Average Score by Test Preparation Course

```{r plot-boxplot-prep}
ggplot(students, aes(x = test_prep, y = average_score, fill = test_prep)) +
  geom_boxplot() +
  labs(title = "Average Score by Test Preparation Course",
       x = "Test Preparation Course", y = "Average Score") +
  theme_minimal() +
  theme(legend.position = "none")
```

**Interpretation:** The boxplot shows both a higher median and a higher
lower-quartile score for students who completed the preparation course,
visually confirming the statistically significant difference found in the
t-test above.

## Reading vs. Writing Score (Relationship)

```{r plot-scatter}
ggplot(students, aes(x = reading_score, y = writing_score)) +
  geom_point(alpha = 0.4, color = "#55A868") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(title = "Relationship Between Reading and Writing Scores",
       x = "Reading Score", y = "Writing Score") +
  theme_minimal()
```

**Interpretation:** Points cluster tightly around the upward-sloping
trend line, visually confirming the strong positive correlation between
reading and writing scores calculated earlier.

## Average Score by Parental Education Level

```{r plot-bar-education}
students %>%
  group_by(parent_education) %>%
  summarise(mean_score = mean(average_score)) %>%
  ggplot(aes(x = parent_education, y = mean_score, fill = parent_education)) +
  geom_col() +
  labs(title = "Average Score by Parental Education Level",
       x = "Parental Education Level", y = "Average Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1), legend.position = "none")
```

**Interpretation:** Average scores tend to rise with parental education
level, with students whose parents hold a master's degree scoring highest
on average and those whose parents completed only some high school
scoring lowest — suggesting a link between parental educational
background and student performance.

## Pass/Fail Proportions by Lunch Type

```{r plot-pass-fail}
ggplot(students, aes(x = lunch, fill = pass_fail)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Pass/Fail Proportion by Lunch Type",
       x = "Lunch Type", y = "Proportion of Students", fill = "Result") +
  theme_minimal()
```

**Interpretation:** A visibly larger share of "free/reduced" lunch
students fall into the "Fail" category (average score below 60) compared
to "standard" lunch students, reinforcing the socio-economic pattern
observed earlier.

# Findings and Conclusion

Based on the import, cleaning, statistical analysis, and visualizations
above, this project finds:

1. **Test preparation matters.** Students who completed the test
   preparation course scored significantly higher on average
   (statistically confirmed with a t-test, p < 0.05).
2. **Socio-economic status is associated with performance.** Students on
   "standard" lunch (a proxy for higher income) outperformed those on
   "free/reduced" lunch, both in average score and in pass rate.
3. **Parental education level correlates positively with student
   scores** — average scores generally increase as parental education
   level rises.
4. **Subject scores are highly inter-correlated.** Math, reading, and
   writing scores are all strongly positively correlated (r > 0.8),
   and reading score alone is a strong linear predictor of writing
   score.
5. **Gender differences** in math score were tested directly; refer to
   the t-test output above for whether the difference in this sample is
   statistically significant.

**Overall conclusion:** Exam performance in this dataset is not driven by
any single factor in isolation. Preparatory effort (test-prep course
completion) and socio-economic/background factors (lunch type, parental
education) all show measurable, statistically supported associations with
student outcomes. This suggests that interventions such as expanding
access to test-preparation resources and support programs targeted at
lower-income students could meaningfully improve academic outcomes.

*Data source: "Students Performance in Exams" dataset (publicly available,
1,000 student records, 8 attributes).*
